Functional pant indicators - wetland, indicator documentation

### 2023-05-09 DRAFT_functional_plant_indicators_seminat.Rmd
introducing beta regression, seminat

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r, include = F}
load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")
```

@@

```{r}
head(wetland.ref.cov.val)
```

@@

```{r, include = F}
load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ANO.RData")

load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO.RData")
```

@@

# similarly for ASO
res.seminat.ASO <- results.seminat.ASO[['2-sided']]
colnames(res.seminat.ASO)

@@

# change region names to something R-friendly
reg$region
reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

@@

### Regions
Alternatively we can calculate and show the region-wise means and standard deviations.
```{r}

@@

regnor <- regnor %>%
  mutate(
    Grazing_mowing1.reg.mean = c(mean(res.seminat2$Grazing_mowing1[res.seminat2$region=="Northern Norway"],na.rm=T),
                             mean(res.seminat2$Grazing_mowing1[res.seminat2$region=="Central Norway"],na.rm=T),
                             mean(res.seminat2$Grazing_mowing1[res.seminat2$region=="Eastern Norway"],na.rm=T),
                             mean(res.seminat2$Grazing_mowing1[res.seminat2$region=="Western Norway"],na.rm=T),
                             mean(res.seminat2$Grazing_mowing1[res.seminat2$region=="Southern Norway"],na.rm=T)),
    Grazing_mowing1.reg.sd = c(sd(res.seminat2$Grazing_mowing1[res.seminat2$region=="Northern Norway"],na.rm=T),
                          sd(res.seminat2$Grazing_mowing1[res.seminat2$region=="Central Norway"],na.rm=T),
                          sd(res.seminat2$Grazing_mowing1[res.seminat2$region=="Eastern Norway"],na.rm=T),
                          sd(res.seminat2$Grazing_mowing1[res.seminat2$region=="Western Norway"],na.rm=T),
                          sd(res.seminat2$Grazing_mowing1[res.seminat2$region=="Southern Norway"],na.rm=T)),
             Grazing_mowing1.reg.n = c(nrow(res.seminat2[res.seminat2$region=="Northern Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                          nrow(res.seminat2[res.seminat2$region=="Central Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                          nrow(res.seminat2[res.seminat2$region=="Eastern Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                          nrow(res.seminat2[res.seminat2$region=="Western Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                          nrow(res.seminat2[res.seminat2$region=="Southern Norway" & !is.na(res.seminat2$Grazing_mowing1),]))
    )

# and adding the values for ASO
regnor <- regnor %>%
  mutate(
    Nitrogen2.ASO.reg.mean = c(mean(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Northern Norway"],na.rm=T),
                             mean(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Central Norway"],na.rm=T),
                             mean(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Eastern Norway"],na.rm=T),
                             mean(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Western Norway"],na.rm=T),
                             mean(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Southern Norway"],na.rm=T)),
    Nitrogen2.ASO.reg.sd = c(sd(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Northern Norway"],na.rm=T),
                          sd(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Central Norway"],na.rm=T),
                          sd(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Eastern Norway"],na.rm=T),
                          sd(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Western Norway"],na.rm=T),
                          sd(res.seminat.ASO2$Nitrogen2[res.seminat.ASO2$region=="Southern Norway"],na.rm=T)),
             Nitrogen2.ASO.reg.n = c(nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                          nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                          nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                          nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                          nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]))
    )

## scaled value maps for Grazing_mowing1 (lower indicator), ANO
# mean
tm_shape(regnor) +
  tm_polygons(col="Grazing_mowing1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
  tm_text("Grazing_mowing1.reg.n",col="black",bg.color="grey")
```
Mean index value by region for the upper Grazing_mowing indicator (i.e. index shows deviations towards less grazing or mowing). Numbers in grey fields show the number of observations in the respective region.

```{r}
# sd
tm_shape(regnor) +
  tm_polygons(col="Grazing_mowing1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("Grazing_mowing1.reg.n",col="black",bg.color="grey")

```
Standard deviation to the mean index value by region for the lower Grazing_mowing indicator. Numbers in grey fields show the number of observations in the respective region.

@@

```{r}
# sd
tm_shape(regnor) +
  tm_polygons(col="Nitrogen2.ASO.reg.sd", title="Nitrogen (upper), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("Nitrogen2.ASO.reg.n",col="black",bg.color="grey")

```
Standard deviation to the mean index value by region for the upper Nitrogen indicator. Numbers in grey fields show the number of observations in the respective region.

@@

Grazing_mowing, ANO:
```{r}
summary(res.seminat$kartleggingsenhet_1m2)
length(unique(res.seminat$kartleggingsenhet_1m2))
# 19 NiN-types, of which T41-C-1 has 0 observations, and T32 & T34 as main types don't have a reference -> so, 16 NiN-Types to plot
colnames(seminat.ref.cov[['Light']])

### Grazing_mowing
par(mfrow=c(4,4))
## T32s
# T32-C-1
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C1C2"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C1',xlab='Grazing_mowing value')
points(density(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-1",]$original,
       rep(0,length(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-1",]$original)),
       col="red")

@@

# T32-C-15
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C15")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C15',xlab='Grazing_mowing value')
#points(density(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-15",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-15",]$original,
       rep(0,length(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-15",]$original)),
       col="red")

@@

summary(res.seminat.ASO$NiN_grunntype)
length(levels(res.seminat.ASO$NiN_grunntype))
# 12 NiN-types, of which C-13 and C-2 have 0 observations -> so, 10 NiN-Types to plot
colnames(seminat.ref.cov[['Nitrogen']])

@@

```
THe figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests. According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.

>>>>

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(readxl)
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r, include = F}
load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")
```

@@

```{r}
head(seminat.ref.cov.val)
```

@@

```{r, include = F}
load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ANO.RData")

load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO.RData")
```

@@

# similarly for ASO
res.seminat.ASO <- results.seminat.ASO[['2-sided']]
#colnames(res.seminat.ASO)

@@

# change region names to something R-friendly
#reg$region
reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

@@

### Regions
Alternatively we can calculate and show the region-wise means and their related standard errors. But note that calculating a simple mean would be inappropriate for these data. This is because 
(i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
(ii) both the ANO and the ASO datasets have a nested structure

Therefore, we need to (i) use a beta-model, that (ii) can account for the nested structure of the data.
Here, we apply the following function using either glmmTMB null-model with a beta-distribution, logit link, and the nesting as a random intercept, or a simple betareg null-model with logit link if the nesting is not extensive enough for a mixed model.
```{r}
library(betareg)
library(glmmTMB)

expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale

# the function performs a glmmTMB if there's >= 5 random levels in the nesting structure
# if that is not the case, then the function performs a betareg if theres >= 2 observations
# if that is not the case either, then the function returns the value of the single observation with NA standard error

indmean.beta <- function(df) {

  st_geometry(df) <- NULL
  colnames(df) <- c("y","ran")

  if ( nrow(df[!is.na(df[,1]),]) >= 2 ) {

    if ( length(unique(df[!is.na(df[,1]),2])) >=5 ) {

      mod1 <- glmmTMB(y ~ 1 +(1|ran), family=beta_family(), data=df)

      return(c(
        expit(summary( mod1 )$coefficients$cond[1]),
        expit( summary( mod1 )$coefficients$cond[1] + 
                 summary( mod1 )$coefficients$cond[2] )-
          expit( summary( mod1 )$coefficients$cond[1] ),
        nrow(df[!is.na(df$y),])
      ))

    } else {

      mod2 <- betareg(y ~ 1, data=df)

      return(c(
        expit(summary( mod2 )$coefficients$mean[1]),
        expit( summary( mod2 )$coefficients$mean[1] + 
                 summary( mod2 )$coefficients$mean[2] )-
          expit( summary( mod2 )$coefficients$mean[1] ),
        nrow(df[!is.na(df$y),])
      ))

  } else {

    return(c(df$y,NA,1))

```{r}

@@

regnor <- regnor %>%
  mutate(
    Grazing_mowing1.reg.mean = c(indmean.beta(df=res.seminat2[res.seminat2$region=="Northern Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat2[res.seminat2$region=="Central Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat2[res.seminat2$region=="Eastern Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat2[res.seminat2$region=="Western Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat2[res.seminat2$region=="Southern Norway",c("Grazing_mowing1","ano_flate_id")])[1]
                                 ),
    Grazing_mowing1.reg.se = c(indmean.beta(df=res.seminat2[res.seminat2$region=="Northern Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat2[res.seminat2$region=="Central Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat2[res.seminat2$region=="Eastern Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat2[res.seminat2$region=="Western Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat2[res.seminat2$region=="Southern Norway",c("Grazing_mowing1","ano_flate_id")])[2]
                               ),
    Grazing_mowing1.reg.n = c(nrow(res.seminat2[res.seminat2$region=="Northern Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                              nrow(res.seminat2[res.seminat2$region=="Central Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                              nrow(res.seminat2[res.seminat2$region=="Eastern Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                              nrow(res.seminat2[res.seminat2$region=="Western Norway" & !is.na(res.seminat2$Grazing_mowing1),]),
                              nrow(res.seminat2[res.seminat2$region=="Southern Norway" & !is.na(res.seminat2$Grazing_mowing1),])
                              )
  )

# and adding the values for ASO
regnor <- regnor %>%
  mutate(
    Nitrogen2.ASO.reg.mean = c(indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1]
                               ),
    Nitrogen2.ASO.reg.se = c(indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2]
                             ),
    Nitrogen2.ASO.reg.n = c(nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]))
  )

## scaled value maps for Grazing_mowing1 (lower indicator), ANO
# mean
tm_shape(regnor) +
  tm_polygons(col="Grazing_mowing1.reg.mean", title="Grazing_mowing (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
  tm_text("Grazing_mowing1.reg.n",col="black",bg.color="grey")
```
Mean index value by region for the upper Grazing_mowing indicator (i.e. index shows deviations towards less grazing or mowing). Numbers in grey fields show the number of observations in the respective region.

```{r}
# se
tm_shape(regnor) +
  tm_polygons(col="Grazing_mowing1.reg.se", title="Grazing_mowing (lower), SE", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("Grazing_mowing1.reg.n",col="black",bg.color="grey")

```
Standard error to the mean index value by region for the lower Grazing_mowing indicator. Numbers in grey fields show the number of observations in the respective region.

@@

```{r}
# se
tm_shape(regnor) +
  tm_polygons(col="Nitrogen2.ASO.reg.se", title="Nitrogen (upper), SE", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("Nitrogen2.ASO.reg.n",col="black",bg.color="grey")

```
Standard error to the mean index value by region for the upper Nitrogen indicator. Numbers in grey fields show the number of observations in the respective region.

@@

Grazing_mowing, ANO:
```{r}
#summary(res.seminat$kartleggingsenhet_1m2)
#length(unique(res.seminat$kartleggingsenhet_1m2))
# 19 NiN-types, of which T41-C-1 has 0 observations, and T32 & T34 as main types don't have a reference -> so, 16 NiN-Types to plot
#colnames(seminat.ref.cov[['Light']])

### Grazing_mowing
par(mfrow=c(2,4))
## T32s
# T32-C-1
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C1C2"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C1',xlab='Grazing_mowing value')
points(density(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-1",]$original,
       rep(0,length(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-1",]$original)),
       col="red")

@@

par(mfrow=c(2,4))
# T32-C-15
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C15")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C15',xlab='Grazing_mowing value')
#points(density(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-15",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-15",]$original,
       rep(0,length(res.seminat[res.seminat$fp_ind=="Grazing_mowing1" & res.seminat$kartleggingsenhet_1m2=="T32-C-15",]$original)),
       col="red")

@@

#summary(res.seminat.ASO$NiN_grunntype)
#length(levels(res.seminat.ASO$NiN_grunntype))
# 12 NiN-types, of which C-13 and C-2 have 0 observations -> so, 10 NiN-Types to plot
#colnames(seminat.ref.cov[['Nitrogen']])

@@

```
The figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests. According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.
